---
title: "Homework 5"
author: "Tommy Bolton"
date: "2023-03-20"
output: pdf_document
---

## Problem 1

```{r, comment=NA, echo=FALSE}
MM2.11 <- function(X){
  dimX <- dim(X)
  f <- cbind(rep(1, dimX[1]), X, X[,2]*X[,3])
  return(f)
}
D_crit <- function(X){
  f <- MM2.11(X)
  crit <- det(t(f)%*%f)
  
  return(crit)
}
A_crit <- function(X){
  f <- MM2.11(X)
  if (det(t(f)%*%f) < (.Machine$double.eps)^(0.5)){
    crit <- Inf 
  }
  else{
    crit <- sum(diag(solve(t(f)%*%f)))
  }
  return(crit)
}
X2.11_vec <- c(1, 1, -1, 1, -1, 1, 1, -1, -1, 1, 1, -1, -1, 1, 1, 1, 1, -1,
           -1, -1, -1, 1, -1, 1, 1, 1, 1, -1, -1, 1, 1, -1, 1, 1, 1, 1,
           -1, -1, -1, -1, 1, 1, -1, 1, -1, -1, 1, 1, 1, 1, -1, -1, 1, -1,
           -1, 1, -1, 1, -1, -1, -1, -1, 1, -1, -1, -1, 1, -1, -1, -1, -1, -1)
X2.11 <- matrix(X2.11_vec, nrow=12, ncol=6, byrow = TRUE)
D_author <- D_crit(X2.11)
D_author
```

```{r, comment=NA, echo=FALSE}
A_author <- A_crit(X2.11)
A_author
```

The D-score for this design is 339738624 and the A-score is 0.7083333.

## Problem 2

```{r, comment=NA, echo=FALSE}
CEXCH_D <- function(N, k){
  X_vec <- runif(N*k, -1, 1)
  X_init <- matrix(X_vec, nrow=N, ncol=k, byrow=TRUE)
  X <- X_init
  dimX <- dim(X)
  D <- D_crit(X_init)
  ## 2) CEXCH iteration
  improvement <- TRUE
  # iteration counter
  n.iter <- 0
  while(improvement){
    n.iter <- n.iter + 1
    D.iter <- D_crit(X)
      for (i in 1:dimX[1]){
        for (j in 1:dimX[2]){
          X1 <- X
          X2 <- X
          X3 <- X
          X1[i,j] <- -1
          X2[i,j] <- 1
          D1 <- D_crit(X1)
          D2 <- D_crit(X2)
          D3 <- D_crit(X3)
        if (D2 < D1 & D3 < D1) {
          X <- X1
          D <- D_crit(X1)
        }
        else if (D2 < D3 & D1 < D3){
          X <- X3
          D <- D_crit(X3)
        }
        else if (D3 < D2 & D1 < D2){
          X <- X2
          D <- D_crit(X2)
        }
      }
    }
    if(D == D.iter){
      # if no improvement was made for a full pass through design, stop
      improvement <- FALSE
    }# end CEXCH while loop
  }
  return(list(initDesign = X_init, initD = D_crit(X_init), 
              optDesign = X, optD = D_crit(X),
              niter = n.iter))
}
```


```{r, comment=NA, cache=TRUE, echo=FALSE}
M <- 2000
res.list_D <- vector("list", length = M)
for(i in 1:M){
  res.list_D[[i]] <- CEXCH_D(N = 12, k = 6)
}
saveRDS(res.list_D, "D_.rds")
scores_D <- unlist(lapply(res.list_D, function(L) L$optD))
Eff_D <- scores_D/D_author
```


```{r, comment=NA, cache=TRUE, echo=FALSE}
boxplot(Eff_D, ylab = "D-efficiency", main = "D-efficiencies of 2000 Candidate Designs")
```


```{r, comment=NA, cache=TRUE, echo=FALSE}
best.found_D <- which.max(scores_D)

res.list_D[[best.found_D]]
```

We found one design that had the same score as the author's but the rest of our designs did not live up to the author's.

## Problem 3

```{r, comment=NA, echo=FALSE}
CEXCH_A <- function(N, k){
  X_vec <- runif(N*k, -1, 1)
  X_init <- matrix(X_vec, nrow=N, ncol=k, byrow=TRUE)
  X <- X_init
  dimX <- dim(X)
  A <- A_crit(X_init)
  ## 2) CEXCH iteration
  improvement <- TRUE
  # iteration counter
  n.iter <- 0
  while(improvement){
    n.iter <- n.iter + 1
    A.iter <- A_crit(X)
      for (i in 1:dimX[1]){
        for (j in 1:dimX[2]){
          X1 <- X
          X2 <- X
          X3 <- X
          X1[i,j] <- -1
          X2[i,j] <- 1
          A1 <- A_crit(X1)
          A2 <- A_crit(X2)
          A3 <- A_crit(X3)
        if (A2 > A1 & A3 > A1) {
          X <- X1
          A <- A_crit(X1)
        }
        else if (A2 > A3 & A1 > A3){
          X <- X3
          A <- A_crit(X3)
        }
        else if (A3 > A2 & A1 > A2){
          X <- X2
          A <- A_crit(X2)
        }
      }
    }
    if(A == A.iter){
      # if no improvement was made for a full pass through design, stop
      improvement <- FALSE
    }# end CEXCH while loop
  }
  return(list(initDesign = X_init, initA = A_crit(X_init), 
              optDesign = X, optA = A_crit(X),
              niter = n.iter))
}
```

```{r, comment=NA, echo=FALSE, cache=TRUE}
M <- 2000
res.list_A <- vector("list", length = M)
for(i in 1:M){
  res.list_A[[i]] <- CEXCH_A(N = 12, k = 6)
}
saveRDS(res.list_A, "A_.rds")
scores_A <- unlist(lapply(res.list_A, function(L) L$optA))
Eff_A <- A_author/scores_A
```

```{r, comment=NA, cache=TRUE, echo=FALSE}
boxplot(Eff_A, ylab = "A-efficiency", main = "A-efficiencies of 2000 Candidate Designs")
```

```{r, comment=NA, cache=TRUE, echo=FALSE}
best.found_A <- which.min(scores_A)

res.list_A[[best.found_A]]
```

The highest efficiency I got was 1 (100%), so I was able to match the quality of author's design on the A-criteria.

## Problem 4

```{r, comment=NA, echo=FALSE}
A_opt <- res.list_A[[best.found_A]]
D_opt <- res.list_D[[best.found_D]]
D_opt_X <- D_opt[[3]]
A_opt_X <- A_opt[[3]]
D_eff_Aopt <- D_crit(D_opt_X)/D_crit(A_opt_X)
D_eff_Aopt
```


```{r, comment=NA, echo=FALSE}
A_eff_Dopt <- A_crit(A_opt_X)/A_crit(D_opt_X)
A_eff_Dopt
```

Both of these designs are 100% efficient on the other criteria so neither one seems to suffer.

```{r, comment=NA, cache=TRUE, echo=FALSE}
CEXCH_D_complex <- function(N, k){
  X_vec <- runif(N*k, -1, 1)
  X_init <- matrix(X_vec, nrow=N, ncol=k, byrow=TRUE)
  X <- X_init
  dimX <- dim(X)
  points <- seq(-1, 1, 0.1)
  D <- D_crit(X_init)
  ## 2) CEXCH iteration
  improvement <- TRUE
  # iteration counter
  n.iter <- 0
  while(improvement){
    n.iter <- n.iter + 1
    D.iter <- D_crit(X)
      for (i in 1:dimX[1]){
        for (j in 1:dimX[2]){
          X1 <- X
          X2 <- X
          X3 <- X
          X4 <- X
          X5 <- X
          X6 <- X
          X7 <- X
          X8 <- X
          X9 <- X
          X10 <- X
          X11 <- X
          X12 <- X
          X13 <- X
          X14 <- X
          X15 <- X
          X16 <- X
          X17 <- X
          X18 <- X
          X19 <- X
          X20 <- X
          X21 <- X
          X22 <- X
          X2[i,j] <- points[1]
          X3[i,j] <- points[2]
          X4[i,j] <- points[3]
          X5[i,j] <- points[4]
          X6[i,j] <- points[5]
          X7[i,j] <- points[6]
          X8[i,j] <- points[7]
          X9[i,j] <- points[8]
          X10[i,j] <- points[9]
          X11[i,j] <- points[10]
          X12[i,j] <- points[11]
          X13[i,j] <- points[12]
          X14[i,j] <- points[13]
          X15[i,j] <- points[14]
          X16[i,j] <- points[15]
          X17[i,j] <- points[16]
          X18[i,j] <- points[17]
          X19[i,j] <- points[18]
          X20[i,j] <- points[19]
          X21[i,j] <- points[20]
          X22[i,j] <- points[21]
          D1 <- D_crit(X1)
          D2 <- D_crit(X2)
          D3 <- D_crit(X3)
          D4 <- D_crit(X4)
          D4 <- D_crit(X4)
          D5 <- D_crit(X5)
          D6 <- D_crit(X6)
          D7 <- D_crit(X7)
          D8 <- D_crit(X8)
          D9 <- D_crit(X9)
          D10 <- D_crit(X10)
          D11 <- D_crit(X11)
          D12 <- D_crit(X12)
          D13 <- D_crit(X13)
          D14 <- D_crit(X14)
          D15 <- D_crit(X15)
          D16 <- D_crit(X16)
          D17 <- D_crit(X17)
          D18 <- D_crit(X18)
          D19 <- D_crit(X19)
          D20 <- D_crit(X20)
          D21 <- D_crit(X21)
          D22 <- D_crit(X22)
          designs <- c(D1, D2, D3, D4, D5, D6, D7, D8, D9, D10, D11, D12, D13,
                       D14, D15, D16, D17, D18, D19, D20, D21, D22)
        if (max(designs)==D1) {
          X <- X1
          D <- D_crit(X1)
        }
        else if (max(designs)==D2){
          X <- X2
          D <- D_crit(X2)
        }
        else if (max(designs)==D3){
          X <- X3
          D <- D_crit(X3)
        }
        else if (max(designs)==D4){
          X <- X4
          D <- D_crit(X4)
        }
        else if (max(designs)==D5){
          X <- X5
          D <- D_crit(X5)
        }
        else if (max(designs)==D6){
          X <- X6
          D <- D_crit(X6)
        }
        else if (max(designs)==D7){
          X <- X7
          D <- D_crit(X7)
        }
        else if (max(designs)==D8){
          X <- X8
          D <- D_crit(X8)
        }
        else if (max(designs)==D9){
          X <- X9
          D <- D_crit(X9)
        }
        else if (max(designs)==D10){
          X <- X10
          D <- D_crit(X10)
        }
        else if (max(designs)==D11){
          X <- X11
          D <- D_crit(X11)
        }
        else if (max(designs)==D12){
          X <- X12
          D <- D_crit(X12)
        }
        else if (max(designs)==D13){
          X <- X13
          D <- D_crit(X13)
        }
        else if (max(designs)==D14){
          X <- X14
          D <- D_crit(X14)
        }
        else if (max(designs)==D15){
          X <- X15
          D <- D_crit(X15)
        }
        else if (max(designs)==D16){
          X <- X16
          D <- D_crit(X16)
        }
        else if (max(designs)==D17){
          X <- X17
          D <- D_crit(X17)
        }
        else if (max(designs)==D18){
          X <- X18
          D <- D_crit(X18)
        }
        else if (max(designs)==D19){
          X <- X19
          D <- D_crit(X19)
        }
        else if (max(designs)==D20){
          X <- X20
          D <- D_crit(X20)
        }
        else if (max(designs)==D21){
          X <- X21
          D <- D_crit(X21)
        }
        else if (max(designs)==D22){
          X <- X22
          D <- D_crit(X22)
        }
      }
    }
    if(D == D.iter){
      # if no improvement was made for a full pass through design, stop
      improvement <- FALSE
    }# end CEXCH while loop
  }
  return(list(initDesign = X_init, initD = D_crit(X_init), 
              optDesign = X, optD = D_crit(X),
              niter = n.iter))
}
```


```{r, comment=NA, echo=FALSE, cache=TRUE}
M <- 2000
res.list_D_complex <- vector("list", length = M)
for(i in 1:M){
  res.list_D_complex[[i]] <- CEXCH_D_complex(N = 12, k = 6)
}
saveRDS(res.list_D_complex, "D_complex.rds")
scores_D_complex <- unlist(lapply(res.list_D_complex , function(L) L$optD))
Eff_D_complex <- scores_D_complex/D_author
```

```{r, comment=NA, cache=TRUE}
boxplot(Eff_D_complex, ylab = "D-efficiency", main = "D-efficiencies of 2000 Candidate Designs")
```

```{r, comment=NA, cache=TRUE, echo=FALSE}
best.found_D_complex <- which.max(scores_D_complex)

res.list_D_complex[[best.found_D_complex]]
```

```{r, comment=NA, echo=FALSE}
max(Eff_D_complex)
```

The best efficiency we found with the new algorithm matches that of our optimal designs before.

```{r, comment=NA, echo=FALSE}
D_opt_complex <- res.list_D_complex[[best.found_D_complex]]
D_opt_X_complex <- D_opt_complex[[3]]
A_eff_Dcomplex <- A_author/A_crit(D_opt_X_complex)
A_eff_Dcomplex
```

This new D optimal design also has an A efficiecny of 1.

```{r, comment=NA, echo=FALSE}
CEXCH_A_complex <- function(N, k){
  X_vec <- runif(N*k, -1, 1)
  X_init <- matrix(X_vec, nrow=N, ncol=k, byrow=TRUE)
  X <- X_init
  dimX <- dim(X)
  points <- seq(-1, 1, 0.1)
  A <- A_crit(X_init)
  ## 2) CEXCH iteration
  improvement <- TRUE
  # iteration counter
  n.iter <- 0
  while(improvement){
    n.iter <- n.iter + 1
    A.iter <- A_crit(X)
      for (i in 1:dimX[1]){
        for (j in 1:dimX[2]){
          X1 <- X
          X2 <- X
          X3 <- X
          X4 <- X
          X5 <- X
          X6 <- X
          X7 <- X
          X8 <- X
          X9 <- X
          X10 <- X
          X11 <- X
          X12 <- X
          X13 <- X
          X14 <- X
          X15 <- X
          X16 <- X
          X17 <- X
          X18 <- X
          X19 <- X
          X20 <- X
          X21 <- X
          X22 <- X
          X2[i,j] <- points[1]
          X3[i,j] <- points[2]
          X4[i,j] <- points[3]
          X5[i,j] <- points[4]
          X6[i,j] <- points[5]
          X7[i,j] <- points[6]
          X8[i,j] <- points[7]
          X9[i,j] <- points[8]
          X10[i,j] <- points[9]
          X11[i,j] <- points[10]
          X12[i,j] <- points[11]
          X13[i,j] <- points[12]
          X14[i,j] <- points[13]
          X15[i,j] <- points[14]
          X16[i,j] <- points[15]
          X17[i,j] <- points[16]
          X18[i,j] <- points[17]
          X19[i,j] <- points[18]
          X20[i,j] <- points[19]
          X21[i,j] <- points[20]
          X22[i,j] <- points[21]
          A1 <- A_crit(X1)
          A2 <- A_crit(X2)
          A3 <- A_crit(X3)
          A4 <- A_crit(X4)
          A4 <- A_crit(X4)
          A5 <- A_crit(X5)
          A6 <- A_crit(X6)
          A7 <- A_crit(X7)
          A8 <- A_crit(X8)
          A9 <- A_crit(X9)
          A10 <- A_crit(X10)
          A11 <- A_crit(X11)
          A12 <- A_crit(X12)
          A13 <- A_crit(X13)
          A14 <- A_crit(X14)
          A15 <- A_crit(X15)
          A16 <- A_crit(X16)
          A17 <- A_crit(X17)
          A18 <- A_crit(X18)
          A19 <- A_crit(X19)
          A20 <- A_crit(X20)
          A21 <- A_crit(X21)
          A22 <- A_crit(X22)
          designs <- c(A1, A2, A3, A4, A5, A6, A7, A8, A9, A10, A11, A12, A13,
                       A14, A15, A16, A17, A18, A19, A20, A21, A22)
        if (min(designs)==A1) {
          X <- X1
          A <- A_crit(X1)
        }
        else if (min(designs)==A2){
          X <- X2
          A <- A_crit(X2)
        }
        else if (min(designs)==A3){
          X <- X3
          A <- A_crit(X3)
        }
        else if (min(designs)==A4){
          X <- X4
          A <- A_crit(X4)
        }
        else if (min(designs)==A5){
          X <- X5
          A <- A_crit(X5)
        }
        else if (min(designs)==A6){
          X <- X6
          A <- A_crit(X6)
        }
        else if (min(designs)==A7){
          X <- X7
          A <- A_crit(X7)
        }
        else if (min(designs)==A8){
          X <- X8
          A <- A_crit(X8)
        }
        else if (min(designs)==A9){
          X <- X9
          A <- A_crit(X9)
        }
        else if (min(designs)==A10){
          X <- X10
          A <- A_crit(X10)
        }
        else if (min(designs)==A11){
          X <- X11
          A <- A_crit(X11)
        }
        else if (min(designs)==A12){
          X <- X12
          A <- A_crit(X12)
        }
        else if (min(designs)==A13){
          X <- X13
          A <- A_crit(X13)
        }
        else if (min(designs)==A14){
          X <- X14
          A <- A_crit(X14)
        }
        else if (min(designs)==A15){
          X <- X15
          A <- A_crit(X15)
        }
        else if (min(designs)==A16){
          X <- X16
          A <- A_crit(X16)
        }
        else if (min(designs)==A17){
          X <- X17
          A <- A_crit(X17)
        }
        else if (min(designs)==A18){
          X <- X18
          A <- A_crit(X18)
        }
        else if (min(designs)==A19){
          X <- X19
          A <- A_crit(X19)
        }
        else if (min(designs)==A20){
          X <- X20
          A <- A_crit(X20)
        }
        else if (min(designs)==A21){
          X <- X21
          A <- A_crit(X21)
        }
        else if (max(designs)==A22){
          X <- X22
          A <- A_crit(X22)
        }
      }
    }
    if(A == A.iter){
      # if no improvement was made for a full pass through design, stop
      improvement <- FALSE
    }# end CEXCH while loop
  }
  return(list(initDesign = X_init, initA = A_crit(X_init), 
              optDesign = X, optA = A_crit(X),
              niter = n.iter))
}
```

```{r, comment=NA, echo=FALSE, cache=TRUE}
M <- 2000
res.list_A_complex <- vector("list", length = M)
for(i in 1:M){
  res.list_A_complex[[i]] <- CEXCH_A_complex(N = 12, k = 6)
}
saveRDS(res.list_A_complex, "A_complex.rds")
scores_A_complex <- unlist(lapply(res.list_A_complex , function(L) L$optA))
Eff_A_complex <- A_author/scores_A_complex
```

```{r, comment=NA, echo=FALSE}
boxplot(Eff_A_complex, ylab = "A-efficiency", main = "A-efficiencies of 2000 Candidate Designs")
```

The highest A efficiency we calculated from the 2000 designs from the $\{-1, -0.9, \cdots, 0.9, 1\}$ swaps was 0.7. Seeing as the rest of our designs had efficiency of 1 on the 2 criteria. We won't want to consider any of these designs.

```{r, comment=NA, echo=FALSE}
RPV <- function(X) { 
  pv <-c()
  f <- MM2.11(X)
  for (x in 1:10000) {
    X1_new <-runif(1, min = -1, max = 1)
    X2_new <-runif(1, min = -1, max = 1)
    X3_new <-runif(1, min = -1, max = 1)
    X4_new <-runif(1, min = -1, max = 1)
    X5_new <-runif(1, min = -1, max = 1)
    X6_new <-runif(1, min = -1, max = 1)
    X_new <- cbind(X1_new, X2_new, X3_new, X4_new, X5_new, X6_new)
    F_new <- MM2.11(X_new)
    predVar <- F_new%*%(solve(t(f)%*%f))%*%t(F_new)
    pv <- cbind(pv, predVar)
  }
  return (pv)
}
randPredVarXA <- RPV(A_opt_X)
SrandPredVarXA <- sort(randPredVarXA)
plot(x=c(1:10000)/10000, y=SrandPredVarXA, main = "Competing Designs", xlab="Fraction of Design Space",
     ylab = "Relative Prediction Variance", type = 'l', col='red')
randPredVarXD <- RPV(D_opt_X)
SrandPredVarXD <- sort(randPredVarXD)
lines(x=c(1:10000)/10000, y=SrandPredVarXD,col="green")
randPredVarXD_complex <- RPV(D_opt_X_complex)
SrandPredVarXD_complex <- sort(randPredVarXD_complex)
lines(x=c(1:10000)/10000, y=SrandPredVarXD_complex,col="blue")
```

I was going to add a legend, but here is the thing, the lines on the FDS plots are all on top of each other. These are the 3 optimal designs we computed, we're excluding the design we computed that optimized A using $\{-1, 0.9, \cdots 0.9, 1\}$ swaps since every design had an efficiency of less than 1 (100%). We can see that these designs always were 100% efficient for both criteria and that they have pretty much the same relative prediction variance over the design space. So because of that, I would just go with whatever design is cheapest of the 3.

Here are the designs.

Here is the only A optimal design we computed.

```{r, comment=NA, echo=FALSE}
A_opt_X
```

Here is the D optimal design we computed using only -1 and 1 swaps.
```{r, comment=NA}
D_opt_X
```

And here is the D optimal design we computed using the $\{-1, 0.9, \cdots, 0.9, 1\}$ swaps

```{r, comment=NA, echo=FALSE}
D_opt_X_complex
```


